\scshape Evolutionarily stable strategy analysis and its links to demography and genetics through invasion fitness

### 2023-01-29 manuscript.tex
almost done with CS section

The definition of an ESS is relatively simple and only involves a set of phenotypes, a measure of fitness (the evolutionary measure of payoff), and a stability condition. Suppose that two individuals interact and the phenotypes that characterize their behavior in the interactions are measured with the variables $x$ and $y$. The phenotypes can be drawn from a set of discrete or continuous set of values that represents a single trait like body size or propensity to cooperate. For the purpose of explaining the theory more concretely, assume that individuals can choose one of two possible behaviors in the interaction, $C$ and $D$. The phenotypes $x$ and $y$ measure the probability that individuals in an interaction with those phenotypes choose behavior $C$, and $1-x$ and $1-y$ are the probabilities those individuals choose behavior $D$ instead (probabilistic phenotypes are called \textit{mixed strategies} in game theory \cite{Osborne:Rubinstein:1994}). For example, the behavior $C$ could be cooperation in the interaction and $D$ could be a lack of cooperation. Let $\w(x,y)$ be the fitness that an individual with phenotype or strategy $x$ receives when interacting with an individual with phenotype or strategy $y$. Fitness $\w$ for the moment is simply the survival rate (i.e., viability selection) but as we'll see we can define much more general fitness measures. The stability condition \cite{Maynard-Smith:Price:1973,Maynard-Smith:1974} says that a phenotype $\ess{x}$ is an ESS, or is evolutionarily stable (ES), when no alternative strategy $y$ can improve upon the fitness that $\ess{x}$ receives when interacting with itself:
\begin{subequations}
  \label{eq:ess:def}
\begin{equation}
  \label{eq:ess:def:a}
  \w(\ess{x}, \ess{x}) \ge \w(y, \ess{x})
\end{equation}
for all phenotypes $y$. In the case that the strategy $y$ gets the same fitness interacting with $\ess{x}$ as $\ess{x}$ does with itself (equation \eqref{eq:ess:def:a} holds as an equality) then $\ess{x}$ must do better interacting with $y$ than $y$ does with itself:
\begin{equation}
  \label{eq:ess:def:b}
  \w(\ess{x}, y) > \w(y, y)
\end{equation}
\end{subequations}
The two conditions in \eqref{eq:ess:def} are equivalent to requiring that phenotype $\ess{x}$ receives higher average fitness than any alternative phenotype $y$ when $y$ is rare in the population \cite{Maynard-Smith:1974,Bishop:Cannings:1976}. If condition \eqref{eq:ess:def} holds as a strict inequality when $y \ne \ess{x}$, then $\ess{x}$ is a ``strict'' ESS.

@@

for all possible mutant $A_{\mut}$ that generate mutant phenotype $x_{\mut}$. This condition is in fact the condition for a strict ESS from \eqref{eq:ess:def} and can be used to prove that the following result for large, randomly mating, diploid populations with two-locus genetic systems \cite{Eshel:Feldman:1984,Eshel:1996,Eshel:Feldman:1998}:

@@

\subsection{Assumptions of long-term evolution}

The degree to which the ESS is a reasonable prediction for long-term phenotypic evolution depends on assumptions made above in developing the long-term evolution process. These assumptions are\footnote{For additional discussion of these assumptions, see \cite{Eshel:1996,Hammerstein:1996,Weissing:1996}}:
\begin{enumerate}[nosep]
  \item Phenotypes are probabilities of selecting among \textit{two} behaviors or strategies. \label{assume:twostrat}
  \item Fitness (i.e., payoff) is a linear function of phenotype. \label{assume:linear}
  \item Individuals interact in pairs that form randomly. \label{assume:pairs}
  \item Populations are large enough to ignore stochastic forces including genetic drift. \label{assume:large-pop}
  \item Mutations are infrequent enough so that a new mutation reaches a short-term equilibrium before additional mutations occur. \label{assume:weak-mut}
  \item Phenotypes are determined by two (or more) haploid or diploid loci. \label{assume:ploidy}
\end{enumerate}
Assumption \ref{assume:twostrat} that there are two behaviors or strategies is particularly important. In practice, games like rock paper scissors \cite{Taylor:Jonker:1978,Zeeman:1980,Weissing:1991,Hofbauer:Sigmund:1998} involving three or more strategies may have more complex long-term evolutionary dynamics resulting in equilibria that may or may not be ESSs. Support for this possibility comes from examples of three strategy interactions like rock paper scissors where the ESS strategy is actually an internally \textit{unstable} equilibrium \cites{Friedman:1991,Weissing:1991}. Weissing sees this issue as a potentially serious limitation for using ESS to describe long-term phenotypic evolution \cite{Weissing:1996}.
Nevertheless, much of the mathematical machinery developed for analyzing external stability can be generalized to three or more strategies \cite[e.g.,][]{Eshel:Feldman:1984,Weissing:1996,Eshel:Feldman:1998,Hofbauer:Sigmund:1998}, and ESS is an appropriate determinant for stability for some three strategy population genetic models \cite[e.g., single-locus three-allele models in continuous time][]{Cressman:Hofbauer:1996}. Thus, future theoretical work might yield conditions under which the two strategy assumption can be relaxed.

The linear fitness function assumption \ref{assume:linear} is also important. It arises naturally from assuming that the fitness of an individual using a mixed strategy phenotype is the expected fitness that individual would obtain using pure strategies with a probability given by the mixed strategy. However, phenotypes do not have to be probabilities of using pure strategies and can instead involve continuous quantities like body size or metabolic rate; fitness can easily a nonlinear function of such phenotypes. Studying nonlinear fitness functions, Eshel, Taylor, and others \cite{Eshel:Motro:1981,Eshel:1983,Taylor:1989,Christiansen:1991} showed that evolutionary stability is more complex than simply the ability of a population with mean phenotype $\ess{x}$ to resist invasion by mutant genotypes with phenotypes not equal to $\ess{x}$.

Some of these of these assumptions may be less restrictive than others. For example, assumption \ref{assume:ploidy} may not be very limiting as many organisms are haploid or diploid and ESS models of other genetic systems like haplodiploidy can be studied by adding demographic structure using matrix population models \cite{Taylor:1990,Taylor:Frank:1996,Frank:1998} like the one described below in section \ref{sec:IF:group-structure} on ESS and inclusive fitness. Assumptions \ref{assume:large-pop} and \ref{assume:weak-mut}

While many natural populations deviate significantly from many if not all of the above assumptions, they may not be as restrictive as they seem for a few reasons. First, it is an open theoretical question as to whether the long-term evolution process occurs under more relaxed assumptions. Second, additional assumptions may relax some of the above assumptions in the long-term process. For example, weak selection on phenotypes allows for the analysis of nonlinear function functions and stochastic forces due to finite population effects and reveals that the ESS condition still describes the endpoints of long-term evolution \cite{Rousset:Billiard:2000,Rousset:2004,VanCleve:2015}. Finally, small enough deviations from the above assumptions may only induce small deviations from long-term evolution predictions from ESS analysis. Thus, ESS analysis might still be useful as a first approximation in more complex scenarios outside the above assumptions. The precise limits of the long-term evolution process and ESS analysis as predictive tools for phenotypic evolution remains a fruit area for future  mathematical, computational, and empirical research.

\subsection{Convergence of long-term evolution}

%% do I put the convergence issue here? maybe in discussion!

@@

It should already be evident given Result \ref{res:twolocus} that the unbeatable phenotypes obtained via external stability are closely connected to phenotypes obtained via an ESS analysis using condition \eqref{eq:ess:def}. In the general case for an arbitrary genetic and demographic scenario, we assume only that we have the invasion fitness $\eig(y, \ess{x})$ of a rare mutant $y$ in the population that otherwise has mean phenotype $\ess{x}$. Starting with the condition for a strict ESS, $w(\ess{x}, \ess{x}) > w(y, \ess{x})$ for $y \ne \ess{x}$ from \eqref{eq:ess:def:a}, we can convert that condition to one in terms of ``relative fitness'' by dividing by the mean population fitness $w(\ess{x}, \ess{x})$ to obtain
\begin{equation}
  \label{eq:strict:relative}
  \frac{w(y, \ess{x})}{w(\ess{x}, \ess{x})} = \omega(y, \ess{x}) < 1 \qquad \text{for all } y \ne x
\end{equation}
where $\omega(y, \ess{x})$ is the relative fitness of a rare mutant population with mean phenotype $y$ in a population with mean phenotype $\ess{x}$. The strict ESS condition in \eqref{eq:strict:relative} using relative fitness is identical to the condition for uninvadability in condition \eqref{eq:unbeatable} using invasion fitness, which means that unbeatable or uninvadable phentoypes are strict ESSs\footnote{The relationship between invasion fitness and non-strict ESSs is more complex and depends on specific details of the genetics of the population. For example, see the section entitled ``Long-term Stability of an ESS'' in \cite{Eshel:Feldman:1998}}.

@@

@@

J.V.C. would like to acknowledge the hard work and extraordinary patience of the editors and staff who brought this special issue together.

>>>>

The definition of an ESS is relatively simple and only involves a set of phenotypes, a measure of fitness (the evolutionary measure of payoff), and a stability condition. Suppose that two individuals interact and the phenotypes that characterize their behavior in the interactions are measured with the variables $x$ and $y$. The phenotypes can be drawn from a set of discrete or continuous set of values that represents a single trait like body size or propensity to cooperate. For the purpose of explaining the theory more concretely, assume that individuals can choose one of two possible behaviors in the interaction, $C$ and $D$. The phenotypes $x$ and $y$ measure the probability that individuals in an interaction with those phenotypes choose behavior $C$, and $1-x$ and $1-y$ are the probabilities those individuals choose behavior $D$ instead (probabilistic phenotypes are called \textit{mixed strategies} in game theory \cite{Osborne:Rubinstein:1994}). For example, the behavior $C$ could be cooperation in the interaction and $D$ could be a lack of cooperation. Let $\w(x,y)$ be the fitness that an individual with phenotype or strategy $x$ receives when interacting with an individual with phenotype or strategy $y$. Fitness $\w$ for the moment is simply the survival rate (i.e., viability selection) but as we'll see we can define much more general fitness measures. The stability condition \cite{Maynard-Smith:Price:1973,Maynard-Smith:1974} says that a phenotype $\ess{x}$ is an ESS, or is evolutionarily stable (ES), when one of two scenarios occurs. For any alternative strategy $y$, either $y$ receives less fitness when interacting with $\ess{x}$ than $\ess{x}$ receives when interacting with itself,
\begin{subequations}
  \label{eq:ess:def}
\begin{equation}
  \label{eq:ess:def:a}
  \w(\ess{x}, \ess{x}) > \w(y, \ess{x}) \, ,
\end{equation}
or $y$ gets the same fitness interacting with $\ess{x}$ as $\ess{x}$ does with itself and $\ess{x}$ receives more fitness interacting with $y$ than $y$ does with itself,
\begin{equation}
  \label{eq:ess:def:b}
  \w(\ess{x}, y) > \w(y, y) \quad \text{when} \quad  \w(\ess{x}, \ess{x}) = \w(y, \ess{x}) \, .
\end{equation}
\end{subequations}
The two conditions in \eqref{eq:ess:def} are together equivalent to requiring that phenotype $\ess{x}$ receives higher average fitness than any alternative phenotype $y$ when $y$ is rare in the population \cite{Maynard-Smith:1974,Bishop:Cannings:1976}. The phenotype $\ess{x}$ is called a ``strict'' ESS when condition \eqref{eq:ess:def:a} holds and a ``weak ESS'' when condition \eqref{eq:ess:def:b} holds.

@@

for all possible mutant $A_{\mut}$ that generate mutant phenotype $x_{\mut}$. This condition is in fact the condition for a strict ESS from \eqref{eq:ess:def:a} and can be used to prove that the following result for large, randomly mating, diploid populations with two-locus genetic systems \cite{Eshel:Feldman:1984,Eshel:1996,Eshel:Feldman:1998}:

@@

@@

It should already be evident given Result \ref{res:twolocus} that the unbeatable phenotypes obtained via external stability are closely connected to phenotypes obtained via an ESS analysis using condition \eqref{eq:ess:def}. In the general case for an arbitrary genetic and demographic scenario, we assume only that we have the invasion fitness $\eig(y, \ess{x})$ of a rare mutant $y$ in the population that otherwise has mean phenotype $\ess{x}$. Starting with the condition for a strict ESS, $w(\ess{x}, \ess{x}) > w(y, \ess{x})$ for $y \ne \ess{x}$ from \eqref{eq:ess:def:a}, we can convert that condition to one in terms of ``relative fitness'' by dividing by the mean population fitness $w(\ess{x}, \ess{x})$ to obtain
\begin{equation}
  \label{eq:strict:relative}
  \frac{w(y, \ess{x})}{w(\ess{x}, \ess{x})} = \omega(y, \ess{x}) < 1 \qquad \text{for all } y \ne x
\end{equation}
where $\omega(y, \ess{x})$ is the relative fitness of a rare mutant population with mean phenotype $y$ in a population with mean phenotype $\ess{x}$. The strict ESS condition in \eqref{eq:strict:relative} using relative fitness is identical to the condition for uninvadability in condition \eqref{eq:unbeatable} using invasion fitness, which means that unbeatable or uninvadable phenotypes are strict ESSs\footnote{The relationship between invasion fitness and non-strict ESSs is more complex and depends on specific details of the genetics of the population. For example, see the section entitled ``Long-term Stability of an ESS'' in \cite{Eshel:Feldman:1998}}.

@@

\section{Assumptions of long-term evolution and ESS}

The previous sections show how ESS analysis using invasion fitness can be used to predict the long-term outcomes of phenotypic evolution in scenarios with both genetic and demographic structure. However, the generality of these results ultimately depends on the assumptions made in developing Result \ref{res:twolocus} in section \ref{sec:long-term}, which relates the ESS condition to equilibria of the long-term evolution process. These assumptions are\footnote{For additional discussion of these assumptions, see \cite{Eshel:1996,Hammerstein:1996,Weissing:1996}}:
\begin{enumerate}[nosep]
  \item Phenotypes are probabilities of selecting among \textit{two} behaviors or strategies. \label{assume:twostrat}
  \item Fitness (i.e., payoff) is a linear function of phenotype. \label{assume:linear}
  \item Individuals interact in pairs that form randomly. \label{assume:pairwise}
  \item Populations are large enough to ignore stochastic forces including genetic drift. \label{assume:large-pop}
  \item Mutations are infrequent enough so that a new mutation reaches a short-term equilibrium before additional mutations occur. \label{assume:weak-mut}
  \item Phenotypes are determined by two (or more) haploid or diploid loci. \label{assume:ploidy}
\end{enumerate}
Some of these of these assumptions are likely less restrictive than others. For example, assumption \ref{assume:ploidy} may not be very limiting as many organisms are haploid or diploid and ESS models of other genetic systems like haplodiploidy can be studied by adding demographic structure using matrix population models \cite{Taylor:1990,Taylor:Frank:1996,Frank:1998} like the one described below in section \ref{sec:IF:group-structure} on ESS and inclusive fitness. The large population size assumption \ref{assume:large-pop} also can be relaxed by using stochastic population models that allow for genetic drift and and use fixation probability as a fitness measure instead of invasion fitness \cite{Rousset:Billiard:2000,Rousset:2003,Nowak:Sasaki:2004,Lessard:2005,VanCleve:2015}. Assumption \ref{assume:weak-mut} is more difficult to relax since doing so allows multiple mutants to occur at a time. Analysis of such multi-mutant invasions is beyond the scope of the mathematical tools typically used in evolutionary analysis. One potential route forward for such analyses might be to follow Luo and Cooney who used non-local integro-differential partial differential equations to study the evolution of cooperation via multilevel selection \cite{Luo:2014,Luo:Mattingly:2017,Cooney:2019,Cooney:2020,Cooney:Mori:2022}.

Assumption \ref{assume:twostrat} that there are two behaviors or strategies is particularly important. In practice, games like rock paper scissors \cite{Taylor:Jonker:1978,Zeeman:1980,Weissing:1991,Hofbauer:Sigmund:1998} involving three or more strategies may have more complex long-term evolutionary dynamics resulting in equilibria that may or may not be ESSs. Support for this possibility comes from examples of three strategy interactions like rock paper scissors where the ESS strategy is actually an internally \textit{unstable} equilibrium \cites{Friedman:1991,Weissing:1991}. Weissing sees this issue as a potentially serious limitation for using ESS to describe long-term phenotypic evolution \cite{Weissing:1996}.
Interestingly, much of the mathematical machinery developed for analyzing external stability can be generalized to three or more strategies \cite[e.g.,][]{Eshel:Feldman:1984,Weissing:1996,Eshel:Feldman:1998,Hofbauer:Sigmund:1998}, and ESS is an appropriate determinant for stability for some three strategy population genetic models \cite[e.g., single-locus three-allele models in continuous time][]{Cressman:Hofbauer:1996}. Thus, future theoretical work might yield conditions under which the two strategy assumption can be relaxed.

The linear fitness function assumption \ref{assume:linear} is also important restriction. It arises naturally from assuming that the fitness of an individual using a mixed strategy phenotype is the expected fitness that individual would obtain using pure strategies with probabilities given by the mixed strategy. Linear fitness functions are convenient in part because they simplify interactions involving more than two individuals: fitness in such interactions can be expressed as a linear combination of the fitness of pairwise interactions. This means the pairwise interaction assumption \ref{assume:pairwise} is unnecessary for linear fitness functions. However, phenotypes often involve continuous quantities like body size or metabolic rate instead of pure strategy probabilities and nonlinear fitness functions are the norm in these cases. Studying nonlinear fitness functions, Eshel, Taylor, and others \cite{Eshel:Motro:1981,Eshel:1983,Taylor:1989,Christiansen:1991,Metz:Geritz:1996,Eshel:Motro:1997,Geritz:Kisdi:1998} showed that evolutionary stability involves two qualitatively different kinds of stability. The first kind is the ability of a population with mean phenotype $\ess{x}$ to resist invasion by any mutant genotypes with phenotype not equal to $\ess{x}$, which is the condition captured by the strict ESS condition \eqref{eq:ess:def:a}. This is the classic notion of evolutionary stability evoked by verbal descriptions of the ESS concept. The second kind of stability is called ``convergence stability'' \cite{Christiansen:1991,Eshel:1996,Geritz:Kisdi:1998}.

\section{Convergence of long-term evolution and ESS}

Convergence stability refers to populations with mean phenotype $x$ near an ESS $\ess{x}$ that are invaded by mutants that bring the population closer to $\ess{x}$ and resist invasion by mutants that move the population farther from $\ess{x}$ \cite{Eshel:1983,Taylor:1989}. Thus, populations near a convergence stable phenotype $\ess{x}$ will evolve towards $\ess{x}$. If an ESS $\ess{x}$ is to really represent the outcome of a long-term evolution process, then it should be convergence stable as well as an ESS. The evolutionary and convergence stability conditions for continuous phenotypes and nonlinear fitness functions are expressed in terms of partial derivatives of the fitness function with respect to the mutant and resident phenotypes \cite{Eshel:1983,Taylor:1989,Geritz:Kisdi:1998}. However, we can still get a sense of a formal statement of the convergence stability condition without the partial derivative expressions by examining the the first inequality in the weak ESS condition \eqref{eq:ess:def:b}, $\w(\ess{x}, y) > \w(y, y)$, which says that $\ess{x}$ should invade a population with an alternative phenotype $y$. In a sense then, the original Maynard Smith and Price definition of an ESS \cite{Maynard-Smith:Price:1973,Maynard-Smith:1974} contained the notion of convergence stability but did not distinguish it from evolutionary stability more generally. Only by examining continuous phenotypes and nonlinear fitness functions did convergence stability emerge clearly as an essential and distinct condition.

Given there are two kinds of stability to asses, evolutionary stability \textit{sensu} condition \eqref{eq:ess:def:a} and convergence stability, a phenotype $\ess{x}$ that is not long-term stable might fail either one of them. If $\ess{x}$ is not convergence stable, then a population with mean phenotype $\ess{x}$ will resist invasion by mutants with alternative phenotypes but populations with mean phenotypes near $\ess{x}$ will eventually evolve away from $\ess{x}$. Such ESSs are unattainable under long-term evolution and are sometimes called ``Garden of Eden'' phenotypes \cite{Hofbauer:Sigmund:1990}. Even more interesting are the cases where the phenotype $\ess{x}$ is convergence stable but not an ESS. This possibility was was not explored much before

@@

J.V.C. would like to acknowledge the hard work and extraordinary patience of the editors and staff who brought this special issue together. Two reviewers provided thoughtful and expert comments that have improved the manuscript. Erol Akçay provided intellectual and emotional support.
